--- 
title: "A new approach to interspecific synchrony in population ecology and its causes using copulas"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Philip C. Reid, Daniel Reuman"
geometry: "left=1 in,right=1 in,top=1 in,bottom=1 in"

output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_MainText.sty
      
tables: True
link-citations: True
urlcolor : blue
indent : True

csl: TheAmericanNaturalist.csl
bibliography: REF_CIS.bib
---

```{r setup_MainText, echo=F}
library(rmarkdown)
knitr::opts_chunk$set(echo = TRUE, fig.pos = "H")
options(scipen = 1, digits = 5) #This option round all numbers appeared in the inline r code upto 5th digit
seed<-101
```

\noindent \emph{Affiliations:}

\noindent Ghosh: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

\noindent Sheppard: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

\noindent Reid: School of Biological & Marine Sciences, University of Plymouth, Drake Circus, Plymouth PL4 8AA, UK; Continuous Plankton Recorder Survey, The Marine Biological Association, The Laboratory, Citadel Hill, Plymouth PL1 2PB, UK

\noindent Reuman: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66047, USA

<!--\noindent\makebox[\linewidth]{\rule{\columnwidth}{1pt}}
\scriptsize-->

\newpage

\noindent \textbf{ABSTRACT} 
<!--200 words limit : J Am Nat, 350 words for J. animal eco.-->
<!--***DAN: Dan has not edited the abstract yet-->
A recent study [@Ghosh_copula] showed the potential uses of copula methods in examining
"spatial synchrony" between metapopulations. 
We used this copula concept to estimate synchronous and/ or compensatory
responses between different species which shared 
the same environment. We applied a partial-Spearman correlation based approach
to measure left and right tail-dependence
between two variables at their extreme left and higher values, respectively. 
Stronger dependence in either tail of the joint distribution of 
two associated species' time-series from the same 
location leads to corresponding asymmetry in inter-specific relationships. 
With two long-term spatio-temporal datasets (\textit{Ceratium spp.} abundance data and 
aphids first flight data), we showed overall asymmetric tail-dependence is a common feature of 
inter-specific synchrony within a community, when we considered all combinations of species-pairs.
We investigated whether this asymmetry at species level was due to similar 
asymmetric tail-dependence of the species in response to their biotic and/ or abiotic drivers.
We considered the abundance of the generalist predator 
\textit{Calanus finmarchicus} and growing season temperature as two key drivers 
which shaped the phytoplankton community, whereas aphids 
first flight dates were influenced by average winter temperature. With a non-species-specific approach for \textit{Ceratium spp.} and and a species-specific apprroach to aphids, we found 
significant correlation between asymmetric tail-dependence of species-driver copulas
and species-species copulas, which explained the probable causal mechanism 
behind such commonly found asymmetric inter-specific response. 

\noindent \textbf{\textit{Keywords:}} tail dependence, copula, inter-species synchrony, aphids, plankton


<!--\noindent\makebox[\linewidth]{\rule{\columnwidth}{1pt}}
\normalsize-->

\newpage 

# Introduction\label{intro}

All ecologists study relationships between biological and environmental 
variables. But standard methods for studying the association between two variables
provide only a small slice of the information content of the association.
For instance, the two pairs of variables in Fig. \ref{pedagogfig}A, B have identical
Pearson and Spearman correlation coefficients, but nonetheless display
very different patterns of association. The variables of Fig. \ref{pedagogfig}A
(respectively, Fig. \ref{pedagogfig}B) are more strongly related in the left
(respectively, right) portions of their distributions,
a potentially important pattern of dependence
that is not captured by the correlation coefficients. Correlations
are not the only way to study associations,
but they are very commonly used, and other standard methods in ecology
provide a similarly limited amount of information.

Statistical approaches exist, however, that provide a complete description
of the relationship between variables; these approaches are applied
widely in fields such as finance and neuroscience, but not in ecology. 
The approaches are based on the idea of the *copula* 
[@nelsen2006_copula; @joe2014_dependence], which we here
introduce in a manner that follows @Ghosh_copula. The potential
of copulas to improve our understanding of ecology was 
also elaborated by @Ghosh_copula. Copulas separate the 
information content of a bivariate dataset, $(x_t,y_t)$ for $t=1,\ldots,T$, 
into two non-overlapping parts: the information in the marginal
distributions (which is not about the association between the variables)
and the rest of the information (which is solely about the association).
The association information is revealed by the plot of $u_t$
against $v_t$, where $u_t$ is the rank of $x_t$ in the set
$\{ x_1,x_2,\ldots,x_T \}$, divided by $T+1$; and $v_t$ is the rank 
of $y_t$ in the set
$\{ y_1,y_2,\ldots,y_T \}$, also divided by $T+1$. Here the rank of the smallest
element of one of these sets is understood to be $1$ and the rank of the 
largest element is $T$. We refer to the $u_t$ and $v_t$ as *normalized ranks*
of the $x_t$ and $y_t$. We refer to the plot of $v_t$ against $u_t$ as 
the *copula plot* for $y_t$ and $x_t$. For instance, the copula plots for
Fig. \ref{pedagogfig}A,B are in Fig. \ref{pedagogfig}C,D, and show the asymmetric 
associations in the tails.
Ranking makes the marginal distributions uniform, isolating only the information
on association between the variables. @nelsen2006_copula and @joe2014_dependence
provide introductory background on copulas. @sklar1959_theorem
proved a theorem stating that copulas give a unique and complete description
of the association between variables. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, copulas have not been used in ecology, except for a few
studies which used a normal copulas [@Valpine2014; @Anderson2018; @Popovic2018], 
which do not really harness the power 
of the technique, and except for @Ghosh_copula.

A major benefit of the copula approach is that it can effectively reveal
*tail dependence*. Tail dependence, already exemplified in Fig. \ref{pedagogfig},
is associations between the more extreme values of distributions. For two 
positively associated variables, *left-tail* (respectively, *right-tail*) 
*dependence* is stronger association
between values in the left or lower portions (respectively, the right or upper
portions) of the two distributions. 

The study of @Ghosh_copula was a wide-ranging study of the importance, 
causes and consequences of copula structures in associations between
ecologically important variables. One of the main foci of that paper was 
associations between fluctuations through time of 
population density or phenological measurements of the same
species in different locations. This study instead focuses on population
density and phenological measurements of different species in the same location.
@Ghosh_copula studied, for instance, associations between first-flight time
series, for a given species of aphid, measured at different locations in 
the UK; and 
associations between plankton density time series, for a given plankton taxon, 
measured at different locations in seas around the UK. We instead study 
associations between first-flight or population density time series
measured in the same location for different (sympatric) species. Thus,
in contrast with the study of @Ghosh_copula, this study 
is more part of community ecology than of spatial ecology.
Our reasons for this shift are as follows.

First, *synchronous* (positively correlated) 
and *compensatory* (negatively correlated) 
population density dynamics
of different species occupying the same area are longstanding 
topics of concern in community ecology, with important ramifications
for the stability or instability of aggregate community or ecosystem
properties [@raimondo2004interspecific; @kent2007synchrony;
@loreau2008species; @gonzalez2009causes; @jochimsen2013compensatory];
there are reasons to believe copula structure of 
associations in this context will play an important
but unstudied role in understanding these topics.
A major past insight into community dynamics [@gonzalez2009causes]
was that an aggregate 
property of a community, such as its total biomass, can be relatively
stable through time although its constituent parts (population 
biomasses of individual species)
are highly variable, if the parts show compensatory dynamics [@hallett2014]. 
Likewise,
synchrony amplifies community biomass variability because
the concordant variations of species biomass time series reinforce each
other [@Ma2017]. If synchronous fluctuations are right-tail dependent,
then species are highly abundant simulataneously, which may produce 
years of extremely high community biomass. Alternatively, if synchronous
fluctuations are left-tail dependent, species are very scarce simultaneously,
potentially producing years of extremely low community biomass. 
Thus the tail dependence of synchrony, not just the presence and strength
of synchrony, may independently influence temporal variability of 
aggregate community properties. 

Second, studies of the phenology of species interacting in one area
have also played a central role in community ecology, with 
important ramifications for whether and to what extent interactions 
will be modified by climate change [@Durant2007; @Yang2010];
there are reasons to believe copula structure of 
associations between variables in this context may play an
important role here, as well.
As climate changes and phenologies shift, there is the potential for 
phenologies of interacting species to shift differently, disrupting
the interaction [@Thackery2010]. This idea is referred to as the match-mismatch
hypothesis. Even if, for instance, year-to-year fluctuations in the emergence times
of two interacting species are 
highly correlated, if this correlation is principally in the 
right tails of the distributions of possible emergence times, 
so that early emergences of the species are actually uncorrelated, then 
mismatched years are likely to occur, impacting the species. Such 
mismatches will occur, in this example, 
when emergence is early. One potential mechanism by which 
early emergences
may be uncorrelated between species while later emergences remain correlated
is if both species follow the same environmental cue for their 
emergence, but physiological limitations of only one of the species prevent 
emergence before a certain date. Advancing emergence dates of myriad species
make this scenario more plausible.

We here begin exploring whether copula structure may be important for 
studies of synchrony and compensatory dynamics, and for studies
of phenology and the match-mismatch hypothesis. We use a 
56-year dataset of population densities of 4 species of dinoflagellates
from the *Certaium* genus, from 15 locations in the seas around the 
United Kingdom (UK); and a 35-year dataset of annual first-flight dates 
for 20 species of aphid from 10 locations within the UK.
Our aphid and plankton
populatuion and phenology variables were 
almost exclusively positively associated with each other (see \nameref{Results}).
<!--***DAN: make sure you explicitly revisit this in Results-->
Furthermore, the terms left- and right-tail 
dependence, defined above, do not apply to negatively associated 
variables, because the 
negative association means values in the left tail
of one variable are associated with those in the right tail of the other; 
distinct methods are required to study tail dependence and its
asymmetry in negatively associated variables. 
Therefore, we introduce methods and present results in this study
chiefly for the case of positively associated variables, 
returning to the topics of negatively 
associated variables and compensatory population dynamics 
in the Discussion.

In addition to examining whether tail dependence in our data is
asymmetric, we also test for possible causes of such patterns.
One possible mechanism, similar to some of the mechanisms 
explored by @Ghosh_copula, is explained for the *Ceratium* example as
follows. Earlier work showed that average sea surface temperature 
is an important
correlate of phytoplankton abundance in our data 
[e.g., @defriez2016climate; @sheppard2017; @sheppard2019]:
cold water is associated with more phytoplankton, likely because upwelling and 
mixing of the surface and deeper ocean layers bring both 
nutrients and cold water to the photic zone. 
However, if it is the case for a given location 
that very cold water is associated with no more
*Ceratium*, on average, than is moderately cold water, then 
that corresponds to a left-tail-dependent 
positive association between the "coldness" of the surface water (measured, for instance,
by how many degrees colder the water is than average)
and *Ceratium* abundance. If such tail dependence is strong and consistent
across the *Ceratium* species for which we have data, it should
produce positive associations with left-tail dependence between 
the abundance time series of different species. Likewise, 
in locations for which the winter coldness-*Ceratium* abundance association
is shows less left-tail dependence, one should see less
left-tail dependence between different *Ceratium* species.
So tail dependence between species may be inherited from joint tail
dependence of both species on a common environmental driver.
Phytoplankton are also strongly influenced by the abundant generalist copepod 
consumer *Calanus finmarchicus*, so our actual investigation of the mechanism 
proposed here will take into account this influence as well as the 
association with sea surface temperature. For aphid first flight, we examine
the same potential mechanism, but the relevant driver in that case is 
winter temperature.

Thus this paper focuses on whether and why population density 
of phenological time series of co-located species may show asymmetric tail 
dependence in their associations, with a focus on positively associated variables
because positive associations are what occured in the 
available data. We ask these specific questions:
(Q1) do synchronous/positively associated population density or phenological time series
of different co-located species commonly show asymmetric tail dependence in the
association? (Q2) If so, what are the causes of these patterns? We regard our investigation
as a first step toward a better understanding of the potential importance of copula
structure and tail dependence for such central ecological topics as synchrony and 
compensatory dynamics in communities and their influence on community stability;
and the match-mismatch hypothesis in phenology. The Discussion has several 
additional thoughts on next steps toward this goal. The conceptual considerations
introduced above are ample evidence, in our view, of the potential for tail
dependence to make a crucial difference in how ecologists understand these 
important topics in ecology.

<!--old introductory fig., keeping it around for a bit in case we need it
\begin{figure}[!h]
\begin{center}
\includegraphics[width=8 cm]{./Results/fig_intro_A.pdf}
\includegraphics[width=8 cm]{./Results/fig_intro_BCDE.pdf} 
\caption{(A) Pedagogical figure of a copula showing stronger dependence in lower tail ( considered 
to be the points lying on left of solid black or red lines) than in upper tail (considered to be the points lying 
on the right of solid black or blue lines). Asymmetric tail dependence in species-driver (negative 
temperature was considered as a driver for aphid species first flight date) copulas (B, D) caused similar 
asymmetric tail dependence in inter-species copulas (C, E).\label{fig_intro}}
\end{center}
\end{figure}
-->

\begin{figure}[!h]
\begin{center}
\includegraphics[width=10cm]{./Results/pedagog_fig.pdf}
\caption{Pedagogical figure for introducing copulas, tail dependence, and partial Spearman correlation.
(A, B) Two pairs of variables that have identical Pearson (P) and Spearman (S) correlation coefficients
but differ markedly in the nature of the association, panel A showing stronger left- than right-tail dependence
and panel B showing the reverse. (C, D) Copula plots (see Introduction) for panels A and B, respectively. 
(E, F) Graphics supporting the definitions of partial Spearman correlation and our statistic measuring 
asymmetry of tail dependence (see Methods).}\label{pedagogfig}
\end{center}
\end{figure}

<!--***DAN: This pedagogical figure is looking better and better,
but can you please make the following modifications:
1) The label for the diagonal line u_t + v_t = 2u_b in panel E
is closer to the line itself than is the label u_t + v_t = 2l_b,
and closer to the labels that occur inside the axes in panel F. Please
make it the same distance.
2) The label u_t + v_t = 0 in panel F looks fine to me. But the label
u_t + v_t = 2 is not aligned in the same way. Please align it in the same
way, for consistency with the other label.

***Shya: done.
-->

<!--***DAN: Shya, I got rid of the list of figs and list of tables from the
sup mat because they were not necessary - we don't have many figs or tables there.
Currently in the supp mat, the title and author names is on one page by itself 
and then Fig. S1 is on the next page. Can you please make Fig. S1 appear
on the first page, together with the title? Also, please slightly increase the
space in between the title and the author names.
***Shya: done.-->

# Methods\label{M&M}

## Data \label{Data} 

Our population dataset comprised average annual 
abundance estimates for 15 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 4 species from the *Ceratium*
genus of dinoflagelates, and for the generalist consumer copepod species
*C. finmarchicus*, for the 56 years 1958 to 2013.
These data were a subset of a larger dataset covering 22 taxa 
and 26 locations, analyzed by @sheppard2017, @sheppard2019, and @Ghosh_copula.
The locations are $2^\circ$ by $2^\circ$ degree grid cells.
The data were originally obtained from the 
Continuous Plankton Recorder (CPR) dataset, 
now operated and maintained by 
the Marine Biological Association of the United Kingdom. 
Data preprocessing steps were the same as used by @Ghosh_copula. 
*Ceratium* species were extracted in part because they
have an important role in red tides, and also because four 
species were available from the genus (Table \ref{tab_plankton_aphid_info}), 
and we wanted closely related 
species because they may compete with one another and may 
be influenced in similar ways
by environmental variables. The 15 locations we used were 
selected from the 26 locations of the larger dataset 
(Fig. \ref{SM-fig_plankton_map}) as follows. 
First, to reduce the effects of sampling variation on statistical results,
we chose the subset of locations for which at least 36 years
<!--***DAN: why 36? Seems odd as stated-->
of data were available for all species.
Second, for a given location, we excluded *Ceratium* species 
that were undetected for more than 10$\%$ of sampled years at that location. 
Finally, we considered only those locations for which at least two 
*Ceratium* species remained. We also had average growing season 
sea surface temperature for each grid cell and year
[@sheppard2017; @sheppard2019]. Earlier analyses (e.g., @sheppard2019)
demonstrated that sea surface temperature and *C. finmarchicus* 
abundance are important covariates of phytoplankton dynamics 
in United Kingdom (UK) seas, 
though associations between temperature and phytoplankton are
probably due to relationships both these variables have with nutrient 
abundance in surface ocean layers. Sea surface 
temperature data preprocessing was the same as used by @sheppard2017.

Our phenology dataset comprised annual first flight dates
for 20 aphid species (Table \ref{tab_plankton_aphid_info})
from 10 locations across the UK (Fig. \ref{SM-fig_aphid_map}),
spanning the 35 years 1976 to 2010. These data were a subset of a larger
dataset covering 11 locations, analyzed previously by @sheppard2016
and @Ghosh_copula. The data were originally obtained from the 
Rothamsted Insect Survey suction-trap dataset 
[@harrington2014; @bell2015]. Data preprocessing 
was the same as that of [@sheppard2016]. 
Locations were screened, leading to the the removal of one of the original
11 sampling locations, by requiring at least 30 years of data be 
available for all species, again to reduce sampling variation of
statistics. We also had time series of winter average 
temperature for each location and year. The winter temerature for 
year $t$ was the average of December of year $t-1$ to March of year $t$.
Earlier analyses have demonstrated the importance of winter temperature
for aphid first flight date (e.g., @sheppard2016).

## Statistical methods\label{Methods}

Given bivariate data $(x_t,y_t)$ for a set of years, $t$, of size $T$, 
and after computing
normalized ranks $(u_t,v_t)$ as described in the Introduction, tail
dependence and asymmetry of tail dependence were measured using the 
*partial Spearman correlation* of @Ghosh_copula, which we here reintroduce. 
The standard Spearman correlation itself measures association between the variables 
$x_t$ and $y_t$ (or between $u_t$ and $v_t$ - recall the Spearman correlation
is based on ranks, so is the same for both sets of variables);
but Spearman correlation measures only the overall 
association of the samples and cannot tell us how association varies across the distributions
of the variables.
Given two bounds $1\leq l_b < u_b \leq 1$, we define the boundary lines 
$u+v=2l_b$ and $u+v=2u_b$ (Fig. \ref{pedagogfig}E), 
which intersect the unit square on which 
a copula is plotted (Introduction, Fig. \ref{pedagogfig}A-D). The partial Spearman correlation for the
band between these boundaries and within the unit square is
\begin{equation}\label{eq.Cor}
   \cor_{l_b,u_b}(u,v) = \frac{\sum 
   (u_t-\mean(u)) (v_t-\mean(v))}{(T-1)\sqrt{\var(u)\var(v)}}.
\end{equation}
\noindent Here, sample means and sample variances are computed using all $T$ data points, 
but the sum, $\Sigma$, is over only the indices $t$ for which $u_t+v_t > 2l_b$ and 
$u_t+v_t < 2u_b$. The partial Spearman correlation is not defined if there
are no points in the band. For positively associated 
$(u_t,v_t)$, the partial Spearman correlations 
$\cor_{0,b}$ and $\cor_{1-b,1}$ for $b\leq 0.5$ 
measure dependence in the left and right tails, respectively, and can be compared
via a difference, $\cor_{0,b}-\cor_{1-b,1}$, to measure asymmetry of tail 
dependence. Positive values (respectively, negative) of this difference mean greater left-tail
(respectively, right-tail) dependence.
<!--***DAN: make sure you *either* use left or lower-tail throughout, similar for right and upper-->
The sum of $\cor_{0,1/2}$ and 
$\cor_{1/2,1}$ (or the sum of
$\cor_{l_{b_k},u_{b_k}}$ for any other choice of bands 
$(l_{b_k},u_{b_k})$ that partition $(0,1)$) equals the standard 
Spearman correlation, as long as no points happen to lie exactly 
on the bounds. 

For each sampling location, $n$, we computed a matrix, $C^n$, which quantifies
asymmetry of tail dependence between aphid or *Ceratium* time series at $n$. Denote by
$s_i^n(t)$ the aphid first flight date or the *Ceratium* population density 
for sampling location $n$, for the $i^{\text{th}}$ species that was present in the 
cleaned data for location $n$, and for year $t$.
We then defined the matrix $C^n$ 
by defining $C^n(i,j)$ for two aphid or *Ceratium* 
species $i,j$, as follows. $C^n(i,j)$ was not defined, or was defined to
equal the missing-data space holder "NA", if $i=j$; or if the hypothesis
that $s_i^n(t)$ and $s_j^n(t)$ were independent could not be rejected 
($5\%$ level, using the `BiCopIndepTest` in the `VineCopula` package in R);
or if independence was rejected but the Kendall correlation
<!--***Shya : in my code I used Kendall, not Spearman-->
of $s_i^n(t)$ and $s_j^n(t)$ was negative. Otherwise we defined 
$C^n(i,j)=\cor_{0,b}(s_i^n(t),s_j^n(t))-\cor_{1-b,1}(s_i^n(t),s_j^n(t))$,
where the partial Spearman correlations in this expression were computed
over the times, $t$, for which data were available for location $n$.
The entry $C^n(i,j)$ was set to NA if independence of 
$s_i^n(t)$ and $s_j^n(t)$ 
could not be rejected because attempting to quantify
tail dependence (or anything else about dependence) for independent
variables is pointless. $C^n(i,j)$ was set to NA for negatively associated
$s_i^n(t)$ and $s_j^n(t)$ because negative association occurred for only 
one pair of species in one location 
in our data (plankton sampling location 18, species *C. furca* and 
*C. macroceros*, see Results). 
Tail dependence
for negatively associated variables should be studied, and this
topic is revisited in the Discussion,
<!--***DAN: in the Discussion, make sure to talk about how studying asymmetry of 
tail dependence for negative associations is complicated by the fact that
it can depend on the arbitrary ordering of the two variables.-->
but negative associations were too rare in our data to study them. 
The matrix $C^n$ is symmetric. The
value $b=1/3$ was used for plankton locations, whereas $b=1/2$
was used for aphid locations because aphid time series were shorter, and
larger $b$ reduces sampling variation for our statistics [@Ghosh_copula].

We also computed a matrix $D^n$ which quantifies tail dependence between 
aphid or plankton time series and their covariates. 
Denote by $d_k^n(t)$ the value of the $k^{\text{th}}$ covariate 
that operated at sampling location $n$ in year $t$ (winter temperature for an
aphid sampling location, sea surface temperature or *C. finmarchicus* density
for a *Ceratium* location). We then defined $D^n$ by defining 
$D^n(i,k)$ for an aphid or *Ceratium* species $i$ and a covariate $k$,
as follows. $D^n(i,k)$ was not defined, or was set to NA, if 
the hypothesis that $s_i^n(t)$ and $d_k^n(t)$ were independent could not 
be rejected ($5\%$ level, `BiCopIndepTest`). Otherwise, if  $s_i^n(t)$ and 
$d_k^n(t)$ were positively associated (positive Kendall correlation)
<!--***Shya : in my code I used Kendall, not Spearman-->, 
we set $D^n(i,k) = \cor_{0,b}(s_i^n(t),d_k^n(t))-\cor_{1-b,1}(s_i^n(t),d_k^n(t))$;
and if  $s_i^n(t)$ and $d_k^n(t)$ were negatively associated, we set
$D^n(i,k) = \cor_{0,b}(s_i^n(t),-d_k^n(t))-\cor_{1-b,1}(s_i^n(t),-d_k^n(t))$.
For aphid first-flight time series, for which $k$ was always $1$ and 
$d_k^n(t)$ was winter temperature in location $n$, associations between
$s_i^n(t)$ and $d_k^n(t)$ were always negative when they were significant
(see Results). The same was true for *Ceratium* density time series and
sea surface temperature. 
<!--***DAN: make sure to explicitly re-mention these facts in the Results,
when you actually get those results-->
Thus our practice of using $-d_k^n(t)$ was
equivalent, in the case of temperature variables, to using a "coldness"
index such as the number of degrees colder than an average 
or typical reference temperature, in place of temperature. Aphid and plankton data 
were always positively associated with the coldness index when they were 
significantly associated with it. Although *C. finmarchicus* abundance was 
positively associated with *Ceratium* time series in some sampling locations
and negatively associated in others, it always showed the same association
with all *Ceratium* species within a location. Using $-d_k^n(t)$ in place
of $d_k^n(t)$ when negative associations with aphid or plankton data occured 
allowed us to study asymmetry of tail dependence using methods developed with 
positively associated variables in mind. Note that using a similar procedure to 
study pairs of negatively associated biological variables (e.g., two aphid or plankton
time series) would be inappropriate because in that case there is no 
cannonical choice of which variable to take the opposite of. We again used
$b=1/3$ for plankton data and covariates, and $b=1/2$ for aphid data and winter
temperature. For display, we horizontally concatenated the matrices
$C^n$ and $D^n$ and displayed matrix values using color. 

We used the matrices $C^n$ for each sampling location $n$ to answer
Q1 from the \nameref{intro}, as follows. First, we counted the number, 
$N_L^n$, of entries of $C^n$ which were not NA and which were greater than 
$0$. These were the "left-tail dominant" species pairs, i.e., pairs of species 
for which left-tail dependence was greater than right-tail dependence.
We also counted the number, $N_R^n$, of right-tail dominant pairs, for which 
<!--***DAN: Shya, note the change in notation, which will require a minor change 
to Figs 2, 3 and the corresponding sup mat figs, please. I think left- and 
right-tail dependence is slightly better than lower- and -upper-tail dependence,
as terminology, just because if you have to say "lower-tail dependence was lower/less than
upper-tail dependence" it sounds weird and a bit confusing. So I changed it throughout.-->
the corresponding entries of $C^n$ were negative. If $N_L^n$ was substantially
greater than (respectively, substantially less than) $N_R^n$ for a location $n$,
it suggested that left-tail dependence (respectively, right-tail dependence) 
between species in that location was dominant, answering Q1 in the affirmative. 
We also calculated $A_{C,L}^n$, the 
sum of all positive, non-NA entries of $C^n$; $A_{C,R}^n$, the sum
of all negative, non-NA entries of $C^n$; and $A_C^n=A_{C,L}^n+A_{C,R}^n$, a
general measure of asymmetry of tail dependence in location $n$. And we calculated
the normalized quantities $F_{C,L}^n=A_{C,L}^n/(A_{C,L}^n+|A_{C,R}^n|)$
and $F_{C,R}^n=A_{C,R}^n/(A_{C,L}^n+|A_{C,R}^n|)$. 
<!--***DAN: Shya, note the changes in notation, which will require some minor 
changes to some of the figs in both the main text and sup mat, please. Note
also that the old f quantities were always positive (you took an absolute values
in the numerator of one of them), but these new quantities are signed in the
same way as the As. I think this is more consistent. If this requires any 
presentational changes, please make them. I expect any required changes will 
be minor, but let me know if not.-->
Together, these statistics provide an answer to Q1.

We used the matrices $C^n$ and $D^n$ to answer Q2 from the Introduction for the 
*Ceratium* and aphid data as follows. First, we calculated 
$A_D^n$, the sum of all non-NA entries of $D^n$. This was analogous to 
$A_C^n$, but calculated using the matrix $D^n$ instead of the matrix
$C^n$. We then examined whether the values $A_C^n$ and $A_D^n$ were 
correlated across locations, $n$. 
This tests the causal hypothesis in the Introduction because it 
tests whether locations for which
*Ceratium* or aphid time series were more right-tail (respectively, left-tail) dependent 
with environmental covariates also had *Ceratium* or aphid time series that were more
right-tail (respectively, left-tail) dependent with each other. 
Recall that an environmental covariate was reversed (its negative was used)
when it was negatively associated with a *Ceratium* or aphid species, and that
no covariate was ever significantly positively associated with some 
*Ceratium* or aphid species and significantly negatively associated with another
such species in the same location (see Results).
<!--***DAN: make sure you explicitly return to this in the Results-->

We also answered Q2 for the aphid data as follows. Within a location, $n$, for each species,
$i$, we computed the sum $\alpha^n_C(i)$ of all non-NA entries $C^n(i,j)$, for 
$j$ ranging across all species for which we had data. This quantity measures an
average tail dependence of species $i$ with other species in the
same location, with positive values for greater left-tail dependence and 
negative ones for greater right-tail dependence. We then defined
$\alpha^n_D(i)$ as the sum of all non-NA entries $D^n(i,k)$, for $k$ 
ranging across all covariates for which we had data. For 
aphids we only had one covariate, winter temperature, so $\alpha^n_D(i)=D^n(i,k)$ 
for $k=1$ corresponding to winter temperature. We provide the more general
definition of $\alpha^n_D(i)$ that applies when more
covariates were available so the definition can also be considered 
(briefly, see below) for *Ceratium* data. We then examined, for each location, $n$, whether 
$\alpha^n_C(i)$ and $\alpha^n_D(i)$ were correlated across species, $i$.
This tests the causal hypothesis in the Introduction because it 
tests whether aphid species which were more right-tail (respectively, left-tail) 
dependent with environmental covariates (winter temperature) also had 
time series that were more right-tail (respectively, left-tail) dependent with 
the time series of other species in the location. Recall that
winter temperature was always negatively associated with aphid first flight
when it was significantly associated (see Results), 
<!--***DAN: make sure to explicitly return to this in the Results-->
and negative temperature (a coldness index) was used in computing $D^n(i,k)$.
Testing whether $\alpha^n_C(i)$ and $\alpha^n_D(i)$ were correlated across
species, $i$, within a location, $n$, was not practical for *Ceratium*,
because we only had data for at most four *Ceratium* species per sampling
location, an insufficient number to provide much statistical power
in testing for a correlation. 

<!--***DAN: Need to say in the Discussion why the correlation across locations
results may be marginally signif for aphids, which may have to do with the test
being no great for aphids. Here some text, taken from other places, that may be useful
when you come to do that:
-We also tested, 
for the aphid data, whether $A_C^n$ and $A_D^n$ were 
correlated across locations, finding only a marginally significant correlation 
(see Results). But this
test is less suitable for the aphid data than it is for the *Ceratium* data, 
for reasons elaborated in the Discussion.
-This 
approach was better when we examined behaviour of fewer but similar species, like the four \textit{Ceratium spp.} 
of dinoflagellate class.
- Our second approach was more suitable when 
there were several species coming from different genus, so expected to show more variation
in inter-species response, as we will see with aphid (20 species) first flight data. 
-->

# Results\label{Results}

<!--Results for Q1-->
We found asymmetric tail-dependence was very common between *Ceratium* population density
time series from the same location, answering Q1 in the affirmative for *Ceratium*;
for some locations, left-tail dependence between *Ceratium*
species was dominant, and for other locations right-tail dependence was dominant.
Associations between *Ceratium* species were always positive when they were significant,
except for one pair of species in one location (plankton sampling location 18, species *C. furca* and 
*C. macroceros*).
For some locations, the matrix $C^n$ was comprised largely of positive values, 
indicating a preponderance of left-tail dependence between *Ceratium* time 
series for the location (Fig. \ref{fig_CorlmCoru_plankton_map_loc12_26}A).
For other locations, $C^n$ had mostly negative values, indicating a 
preponderance of right-tail dependence (Fig. \ref{fig_CorlmCoru_plankton_map_loc12_26}B).
The statistics $F_{C,L}$ and $F_{C,R}$, plotted across all sampling locations
(Fig. \ref{fig_CorlmCoru_plankton_map_loc12_26}C), indicated that most *Ceratium*
sampling locations were dominated by either left- or right-tail dependence, with
approximately equal numbers of each, with only a few locations having
more symmetric tail dependence, on average across pairs of *Ceratium* species.

<!--***DAN: got to here, 2019 07 15, 1126am-->


We got similar findings for aphid firsflight data (Fig. \ref{fig_CorlmCoru_ff_map_loc2_5}). Proportion of lower 
($f_L$) and upper ($f_U$) tail dependence estimates in overall tail-asymmetry was summarized for 10 selected 
sampling sites in Fig. \ref{fig_CorlmCoru_ff_map_loc2_5}(A). $8$ out of $10$ sites were dominantly upper 
tail dependent in $C_{L-U}^n$ (site codes in blue color of Fig. \ref{fig_CorlmCoru_ff_map_loc2_5}A),
indicating overall synchronous late arrival of 20 aphids at those $8$ locations.
Two representative cases for major lower 
(where $N_L > N_U$ for site $n = 2$) and upper (where $N_U > N_L$ for site $n = 5$) tail 
dependence matrix $C_{L-U}^n$ are pictured in 
Fig. \ref{fig_CorlmCoru_ff_map_loc2_5}(B) and (C), respectively. Major asymmetric tail-dependence 
observed with one set of population data (plankton)
and one set of phenological data (aphids first flight) answered (Q1), indicating that inter-specific 
synchrony is not well described by standard correlation metrics, as these 
obscure the fact that species may be simultaneously rare, or simultaneously 
common, with the same resultant correlation coefficient.

<!--plankton plots : Q1-->
\begin{figure}[!h]
\begin{center}
\textbf{ \hspace{1 cm} (A) \hspace{7 cm} (B)} \\
\includegraphics[width=8 cm]{./Results/plankton_results/npa_stat_results/loc12/loc12_Corl-Coru.pdf}
\includegraphics[width=8 cm]{./Results/plankton_results/npa_stat_results/loc26/loc26_Corl-Coru.pdf}
\textbf{ (C)} \\
\includegraphics[width=8 cm]{./Results/plankton_results/npa_stat_results/Corstat_LmU_values_on_map_sp_only.pdf} \\
\caption{Either right- or left-tail dependence between population density time series of 
\emph{Ceratium} species could dominate, depending on the sampling location. (A, B) The matrices
$C^n$ and $D^n$ (Methods), horizontally concatenated, for example locations $n=12$ (A) and $n=26$ (B).
See Table \ref{tab_plankton_aphid_info} for species names. 
All the non-NA values in $C^n$ were positive (red) for location $12$ (A), indicating left-tail
dependence dominated in that location; but values were largely negative (blue) for location
$26$ (B), indicating right-tail dependence dominated there. Matrix entries which were NA 
because time series were independent are displayed in yellow. The counts $N^n_L$ and $N_n^R$
(see Methods) also reflect the distinct tail dependence characteristics of the two
locations. \emph{C. fin.} = \emph{C. finmarchicus}; Temp. = temperature.
Green dots in $D^n$
represent variables which were originally negatively associated, so the negative
of the environmental covariate was used for calculating tail dependence.
See Fig. \ref{SM-fig_CorlmCoru_plankton_all_loc} 
for analogous figures for the other sampling locations. 
(C) The summary statistics $F_{C,L}$ and $F_{C,R}$ (see Methods) 
for each site show that dependence between \emph{Ceratium} species was either 
substantially dominanted by the left or right tails of \emph{Ceratium} distributions, with the
exceptions of a few locations for which tail dependence was closer to symmetric. Site codes
are colored red or blue depending on which of $F_{C,L}$ or $F_{C,R}$ had higher magnitude. Values are
not plotted for site 3 because the hypothesis could not rejected for that site
that dynamics of distinct \emph{Ceratium} species were independent. 
\label{fig_CorlmCoru_plankton_map_loc12_26}}
\end{center}
\end{figure}
<!--***DAN: Shya, please see if you can make "C.fin." appear italicized on the figure.-->

<!--aphid plots : Q1-->
\begin{figure}[!h]
\begin{center}
\textbf{ (A)} \\
\includegraphics[width=8 cm]{./Results/aphid_results/ff_npa_stat_results/Corstat_LmU_values_on_map_sp_only.pdf} \\
\vspace{1 cm}
\textbf{ (B) \hspace{7 cm} (C)} \\
\includegraphics[width=8 cm]{./Results/aphid_results/ff_npa_stat_results/loc2/loc2_Corl-Coru.pdf}
\includegraphics[width=8 cm]{./Results/aphid_results/ff_npa_stat_results/loc5/loc5_Corl-Coru.pdf}
\caption{$(A)$ Proportion of tail dominant contribution ($F_{C,L}$ for lower tail and $F_{C,R}$ for upper tail 
showing on a color scale between 0 to 1) in overall asymmetry for each of the
selected 10 locations (see Fig. \ref{SM-fig_aphid_map}) summarizes the overall 
asymmetry in tail-dependence for all sampling sites. Sites with lower tail dominance
have site code in red color, whereas upper tail dominant sites have blue site code. 
Lower tail - upper tail dependence matrix 
($C^n$) with first flight data of 20 aphid species are shown in (B) for location n = 2 and
in (C) for location n = 5, for other locations see Fig. \ref{SM-fig_CorlmCoru_ff_all_loc}. 
The last row and column indicates estimates from the interaction between each 
aphid species and winter average temperature (which was supposed to be the driver of the species' synchrony),
details were discussed in \nameref{Results}. $N_L^n$ and $N_R^n$ are the count of positive and 
negative cells for 20$\times$ 20 matrix; $N_L^n > N_R^n$ in $(B)$ 
indicating major lower tail dependence in the sp-sp interaction matrix. Similarly, $N_R^n > N_L^n$ in $(C)$ indicates more upper
tail dependence. Yellow cells indicate independent relationship. Cells with a green dot indicate the associated 
variables were negatively correlated initially and hence we reversed the variables in column so that the copula had positive association.\label{fig_CorlmCoru_ff_map_loc2_5}}
\end{center}
\end{figure}


<!--Results for Q2-->
To explain asymmetric tail-dependence between 
\textit{Ceratium spp.} ($s = 4$), we considered abundance of predator C. fin 
and growing season temperature as two key drivers ($d = 2$) for inter-specific synchrony. 
We observed species synchronous responses towards biotic and/ or abiotic stress
which transmitted similar asymmetric tail-dependence from drivers level ($A_D^n$ or $\alpha_D^n$)
to species level ($A_S^n$ or $\alpha_S^n$). 
We did a two-tailed correlation test between $A_S^n$ and $A_D^n$ for all selected location $n$ 
excluding $n = 3$ (as site 3 was independent). We got significant 
positive Pearson correlation ($0.699$)<!--caution : not automated link--> with p-value $<<0.025$, 
which confirmed our hypothesis for (Q2) that tail-asymmetry translated from driver to species level 
(Fig. \ref{fig_plankton_scatter}). In other words, if all \textit{Ceratium} species were 
particularly correlated in their lower tails with their drivers, they were
particularly correlated in their lower tails with each other, and vice versa. 
For aphids, we considered winter average temperature as 
the main driver $(d = 1)$ for their synchronous first flight date at a given location, $n$. 
We tried a similar approach as of \textit{Ceratium spp.} to explain the causal mechanism for 
asymmetric tail-dependence at species level. Since aphids were 20 different sp. which 
varied widely in their inter-specific response and were also not expected 
to have direct interactions as of the \textit{Ceratium} community,
$A_S^n$ and $A_D^n$ could not illuminate the causal relationship 
(Fig. \ref{SM-fig_aphid_scatter}, non-significant Pearson correlation with $p = 0.0678$). 
In an alternative approach, for each of the selected 
10 locations, we computed the Kendall's correlation coefficient $(K)$ between
$\alpha_S^n(i)$ and $\alpha_D^n(i)$, where
$i = 1, 2, \dots ,20$ represented the species index. We found for the $i^{th}$ aphid species, the 
tail dependence of species-temperature copula results in similar tail dependence in inter-species copula 
for 7 out of 10 sites (based on p-value with $0.05$ significance level, Fig. \ref{fig_aphid_multipanel}),
which answered the probable causal mechanism behind inter-species synchrony (Q2). In other words, 
if an aphid species was particularly correlated in its lower tail with the driver, it 
was particularly correlated in its lower tail with the other aphid species.


<!--plankton scatter plot : Q2-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=10 cm]{./Results/plankton_results/npa_stat_results/Corstat_scatter_LmU_values.pdf} 
\caption{Siginificant positive Pearson correlation (p value based on two-tailed test) between total 
asymmetry on species level $A_C^n$ and on driver level $A_D^n$ represents causal relationship between 
tail-dependence of inter-species synchrony with its drivers (C. fin. and growing season temperature) for 
\textit{Ceratium spp.}, numbers in the plot represent the side code $n$.\label{fig_plankton_scatter}}
\end{center}
\end{figure}


<!--aphid multipanel plot : Q2-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/aphid_results/ff_npa_stat_results/Corstat_LmU_avg_sp_temp_multipanel_plot.pdf} 
\caption{For 7 out of 10 sites, the Kendall correlation between $\alpha_S^n(i)$ and $\alpha_D^n(i) ; i = 1, 2, \dots, 20$, 
is significant (p $< 0.05$, one tailed test). This indicates asymmetric tail-dependence between species-winter
average temperature is the reason for similar tail-dependence between the $i^{th}$ species
and the other species.\label{fig_aphid_multipanel}}
\end{center}
\end{figure}


# Discussion\label{Discussion}

<!--summary results-->
We found  substantial inter-specific synchrony with asymmetric tail-dependence between different 
species sharing the same location through time.  With two long-term spatio-temporal 
datasets from two different taxa, we showed partial-Spearman correlation 
estimates at extreme tails of copulas could be used as a simple tool to study dependence 
interactions between species. We also analyzed the probable causal mechanism of observed asymmetric 
tail-dependence. We hypothesized asymmetric interactions (biotic/ abiotic) of species with its 
environmental drivers which transmitted similar tail-dependence through the inter-species level. We 
verified our hypothesis with two alternative approaches for plankton and aphids, respectively. We think 
our approaches could be useful in investigating inter-species dependence structures for more
complex systems like foodwebs with multiple trophic interactions.

<!--why aphids showing more upper tail dep. like in fig1e, rather than 1c?-->
We found aphids were more common with synchronous late arrival (upper tail dependence;
$N_U > N_L$) at 8 out of 10 sites. This was attributed to the fact that aphid first flight dates
were limited by cold winter effects (Fig. \ref{fig_intro}D), causing a dependence structure like Fig. \ref{fig_intro}E 
more frequently than Fig. \ref{fig_intro}C. We believe this is explained by 
the particular sensitivity of overwintering aphids to frost conditions
(killing non-winged morphs) delaying the population reaching the winged-morph stage, and first flight.

<!--Why consider only postively correlated cells ?-->
We considered only positively correlated species pair in explaining the causal 
relationship of inter-specific synchrony with its drivers. All aphids showed positive 
correlation with each other at all selected locations 
(Figs. \ref{fig_CorlmCoru_ff_map_loc2_5}, \ref{SM-fig_CorlmCoru_ff_all_loc}). 
For \textit{Ceratium} species, species $4$ (\textit{Ceratium macroceros}) was 
found to have no significant correlation with either 
driver at locations $n = 18$ and $n = 21$, unlike the other species which were typically found to be 
negatively correlated with the drivers at locations where data was available 
(Fig. \ref{SM-fig_CorlmCoru_plankton_all_loc}).  Thus we do not expect 
these drivers to influence the behaviour of species $4$ in either the upper or lower tail of its 
distribution, and in fact at location $18$ species $4$ was found to be negatively correlated 
with another \textit{Ceratium} species (sp = 2: \textit{Ceratium furca}), 
unlike all other inter-species correlations which were found to be positive 
due to their common relationship with the drivers. The tail dependence $C_{L-U}^{n}(2,4)$ and
$C_{L-U}^{n}(4,2)$ associated 
with this negative relationship were not included in the calculation 
of asymmetry estimates ($A_S^{n+}$, $A_S^{n-}$, $A_S^{n}$) at location $n=18$, as described in methods, 
and our hypothesis that the inter-species tail dependence measures would match driver-species
tail dependence measures cannot be applied to this species.

<!--what extra info do you get by monitoring tail dep. for plankton?-->
Monitoring tail dependence (a measure of "synchrony") could reveal interesting information about 
an ecosystem. We expect our resuts to generalize to other plankton systems subject to limiting factors.
In most freshwater lakes algal-productivity is limited by the availability 
of inorganic phosphorus, represented by lower tail-dependence (stronger association
between smaller values) in a copula made of plankton abundance and phosphorous level.
But excessive fertilizers (through sewage or agricultural runoff) beyond a threshold 
can catalyze plankton-growth. Phytoplankton bloom, an
ecosystem-level perturbation, occurs when compensatory dynamics between perturbation-sensitive
phytoplankton and perturbation-tolerant taxa become weak and so dramatic changes occur in the total
biomass on aggregate level. As phytoplankton communities are diverse in response to change
in the nutrient level in a lake, monitoring their individual dynamics through copula approaches 
could help to better understand the phytoplankton-resilience on the community level and 
even could be used as an "early warning" before the actual bloom occurs.

<!--plankton regime shift (CALFIN-CALHEL) depending on temperature or salinity?-->

# Acknowledgments

We thank the many contributors to the large datasets we used; D. Stevens and P. Verrier for 
data extraction; and Joel E. Cohen, Lauren Hallet, and Jonathan Walter for helpful 
suggestions.
<!--***DAN: Shya and Lawrence to add any other acknowledgments. Shya, please check with
Lawrenec by emailing or talking to him as he is not likely to read this comment.-->
We thank James Bell of the Rothamsted Insect Survey (RIS). The RIS, a UK Capability,
is funded by the Biotechnology and Biological Sciences Research Council under the Core
Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National Science
Foundation grant 1714195 and the James S McDonnell Foundation.  





<!--Table with species info -->
```{r tab_plankton_aphid_info, echo=F, results='asis',message=F}
library(tinytex)
library(tibble)
library(kableExtra)
library(dplyr)

#aphid_info_org <- tibble(c0=c(1:20),
#                     c1=c("Apple grass aphid","Bird cherry oat aphid","Black bean aphid","Blackberry cereal aphid",
#                       "Blackcurrant sowthistle aphid",
#                      "Corn leaf aphid","Currant lettuce aphid","Damson hop aphid","Grain aphid","Green spruce aphid",
#                      "Leaf-curling plum aphid","Mealy cabbage aphid","Mealy plum aphid","Pea aphid","Peach potato aphid",
#                      "Potato aphid","Rose grain aphid","Shallot aphid","Sycamore aphid","Willow carrot aphid"),
#                    c2=c("Rhopalosiphum insertum", "Rhopalosiphum padi", "Aphis fabae", "Sitobion fragariae", "Hyperomyzus lactucae",
#                      "Rhopalosiphum maidis", "Nasonovia ribisnigri","Phorodon humuli","Sitobion avenae","Elatobium abietinum",
#                       "Brachycaudus helichrysi","Brevicoryne brassicae","Hyalopterus pruni","Acyrthosiphon pisum","Myzus persicae",
#                       "Macrosiphum euphorbiae","Metopolophium dirhodum","Myzus ascalonicus","Drepanosiphum platanoidis","Cavariella #aegopodii")
#)

plankton_aphid_info <- tibble(c0=c(1:4,1:20),
                     c2=c("Ceratium fusus","Ceratium furca","Ceratium tripos","Ceratium macroceros",
                          "Rhopalosiphum insertum", "Rhopalosiphum padi", "Aphis fabae", "Sitobion fragariae",
                          "Hyperomyzus lactucae", "Rhopalosiphum maidis", "Nasonovia ribisnigri","Phorodon humuli",
                          "Sitobion avenae","Elatobium abietinum", "Brachycaudus helichrysi","Brevicoryne brassicae",
                          "Hyalopterus pruni","Acyrthosiphon pisum","Myzus persicae", "Macrosiphum euphorbiae",
                          "Metopolophium dirhodum","Myzus ascalonicus","Drepanosiphum platanoidis","Cavariella aegopodii"                                           ))
knitr::kable(plankton_aphid_info, "latex", booktabs = T, linesep = "",align="c",
      caption = "Species names of 4 plankton and 20 aphids for which data were used. \\label{tab_plankton_aphid_info}",
      col.names = NULL)%>%
      group_rows("Plankton", 1,4) %>%
      group_rows("Aphids", 5,24) %>%
      column_spec(column = 2,italic=T)%>%add_header_above(header=c("Species id" = 1, "Latin binomial" = 1))
```













\clearpage

# Literature cited

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{1pt}
\noindent
