--- 
title: "A new method to estimate interspecific synchrony and its causes in population ecology using ``copula'' approach"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Philip C. Reid, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1 in,right=1 in,top=1 in,bottom=1 in"

output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_MainText.sty
      
tables: True
link-citations: True
urlcolor : blue
indent : True

csl: TheAmericanNaturalist.csl
bibliography: REF_CIS.bib
---

```{r setup_MainText, echo=F}
library(rmarkdown)
knitr::opts_chunk$set(echo = TRUE, fig.pos = "H")
options(scipen = 1, digits = 5) #This option round all numbers appeared in the inline r code upto 5th digit
seed<-101
```

\noindent \emph{Affiliations:}

\noindent Ghosh: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

\noindent Sheppard: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

\noindent Reid: School of Biological & Marine Sciences, University of Plymouth, Drake Circus, Plymouth PL4 8AA, UK; Continuous Plankton Recorder Survey, The Marine Biological Association, The Laboratory, Citadel Hill, Plymouth PL1 2PB, UK

\noindent Reuman: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66047, USA; Laboratory of Populations, Rockefeller University, 1230 York Ave, New York, NY, 10065, USA

<!--\noindent\makebox[\linewidth]{\rule{\columnwidth}{1pt}}
\scriptsize-->

\newpage

\noindent \textbf{ABSTRACT:} 
<!--\small write abstract here-->
<!--200 words limit : J Am Nat-->


\noindent \textbf{\textit{Keywords:}} tail dependence, copula, inter-species synchrony, aphids, planktons


<!--\noindent\makebox[\linewidth]{\rule{\columnwidth}{1pt}}
\normalsize-->

\newpage 

# Introduction\label{intro}
<!--Q1 : Does cross-species synchrony have any specific tail dependence, 
Q2: if yes, then what is the reason behind that?-->

<!--Motivation of the paper-->
Studying spatial patterns of 
metapopulation-abundance through time and identifying 
the cause behind their correlated temporal 
flucuation (known as "spatial synchrony") have been of common
interest to ecologists [@lande1999; @liebhold2004; @walter2017]. 
Likewise, estimating synchronous or compensatory dynamics
among different species occupying a common geographical area
is a longstanding focus in ecology to predict the stability of 
aggregate community or ecosystem properties [@raimondo2004interspecific; @kent2007synchrony;
@loreau2008species; @gonzalez2009causes; @jochimsen2013compensatory]. In general,
past studies used correlation based approach or regression based analysis
to characterize temporal fluctuations of co-occuring species and thereby
overlooked the information of the dependence structure embedded in multi-species 
distribution over time. Our recent study [BIVAN paper] showed 
the concept of "copula" can be useful in quantifying "spatial synchrony" 
among metapopulations and could give us more insight in explaining the 
causes and consequences of their synchronous response. But in case of inter-species 
synchrony, use of copula, to our knowledge, remains unexplored. Here, we will 
demonstrate the scope of using the statistical concept of "copula" in quantifying
the synchronous or compensatory fluctuations of multiple co-occuring species over time.

<!--what is copula?-->
Copulas give a unique and 
complete representation of the dependence between variables, in the sense that
the joint distribution of two variables can be reconstructed 
completely from the copula and the marginal distributions [@sklar1959_theorem]. 
Copula approaches [@joe2014_dependence; @nelsen2006_copula] separate the information contained in a bivariate
random variable into two distinct parts: the information contained in the marginal distributions
(which says nothing about the dependence between the variables), 
and the remaining information (which is solely about the dependence between the
variables). To elaborate, a bivariate copula can be constructed from a set of two variables
$\{ x_i, y_i\}, i= 1,2,...,n$ after some ranking procedure is applied so that their marginal
distributions become uniform. If $\{ \bar{u_i}, \bar{v_i}\}$ are the corresponding ranks 
of $\{ x_i, y_i\}$, then $\{ u_i, v_i\} = \{ \bar{u_i}, \bar{v_i}\} / (n+1)$
reveals the associated dependence in the bivariate copula when plotted in a unit square.
Thus, measures of association which depend only on the copula and not on the marginal
distributions (e.g., Kendall's $\tau$ and Spearman's $\rho$) remain unchanged but usual
Pearson linear product-moment correlations of variables and of their copula transformed 
form are not the same. As a result, correlation coefficients measure only the overall 
association and could not tell us how their association varies across the distribution.
Through suitable choice from a wide varity of families and parameters, 
a good deal of control can be achieved in modelling the dependence structure of a bivariate
copula which shows over what parts of the distributions the 
variables are more strongly associated [REF BIVAN]. One aspect emphasized 
below is estimating the strength of the relationship between variables
in the tails of the distributions. 

<!--what is tail-dep.-->
Tail dependence is association 
between extreme values of two variables in a copula. 
Therefore, if smaller values of two positively associated variables 
are more strongly associated than the larger values, the variables 
are said to have stronger 
lower-tail dependence than upper-tail dependence. Similarly, stronger upper tail 
dependence in copula results from two variables correlated strongly in their higher values.
In either case, those variables give rise to asymmetric copula structure. Several families of copula
are available with positive tail dependence exclusively in either
or both tail [@joe2014_dependence; @nelsen2006_copula], in addition to 
a copula from 'Normal' family, having 
zero tail-dependence (in both lower and upper tail, and hence symmetric copula). Following 
we will discuss the relevance of 
copula with its unique tail-dependence feature in modelling multi-species
ecological response.

<!-- How tail-dep. is relevent in multi-species interaction?-->
Estimating tail-dependence could be
accounted for an alternative metric to measure synchrony (or compensation)
between two correlated ecological variables. For example, 
if we consider a bivariate copula made of 
two co-existing species abundance data (through years),
which are positively correlated, then we can estimate their interspecific 
response by measuring the tail-dependence of the constructed copula. Two 
synchronously rare (or abundant) species will show stronger 
lower-tail (or upper-tail) dependence, 
when their abundance-time series data will be plotted as a bivariate 
copula with positive correlation. But if the two species are competing
with each other for same resources then they are negatively correlated and in that 
case weak competitor species will be limited by the abundance of stronger 
one. Thus, here tail-dependence could also illuminate the causal relationship between
interspecific dynamics. 

<!--Why tail-dep is important?------explain : 1. Leibigs law, 2.for aggregated abundance?-->
Monitoring tail dependence could reveal interesting information about 
an ecosystem. At any given time in a particular ecosystem,
productivity is constrained by a single, metabolically essential factor that is
present in least supply relative to the potential biological demand (Leibig's law).
This limiting factor could be climatic (temperature, windspeed, salinity 
or moisture) or the factor could involve an insufficient
supply of a particular nutrient, or an excessive, toxic availability of another chemical.
In this sense, the limiting factor represents a type of ecological stress which
if alleviated will result in greater productivity and development of the ecosystem.
For example, extreme temperature could have profound effect on plant-productivity 
or its pollination rate depending upon the species [@hatfield2015]. Copula structure with 
lower (or upper) tail-dependence could reveal the scenario whether temperature (or any other
abiotic/ biotic variable) suppress (or accelerate) species' productivity. 
Similarly, in most freshwater lakes algal-productivity is limited by the availability 
of inorganic phosphorus, thus represented by lower tail-dependence (stronger association
between smaller values) in a copula made of plankton abundance and phosphorous level.
But excessive fertilizers (through sewage or 
agricultural runoff) beyond a threshold can catalyze their growth. Phytoplankton bloom, an
ecosystem-level perturbations, occurs when compensatory dynamics between perturbation-sensitive
phytoplankton and perturbation-tolerant taxa 
become weak and thereby affecting the total biomass on aggregate level. As phytoplankton
communities are diverse in response to change in the nutrient level in a lake, monitoring 
their individual dynamics through copula could help to better understand the phytoplankton-resilience
on the community level.

In this paper, we will show, by minimum 
AIC-weight based model selection approach, how to measure 
the "tail-dependence" between inter-species dependence structure.
We will also show an alternative non-parametric analysis to verify
model selection based findings. Here we will analyze inter-species synchrony 
with two multi-species
spatio-temporal datasets of two different taxa 
(aphid and plankton) to answer the following questions: 
(Q1) Do different species show synchronous behavior through asymmetric tail 
dependence (particularly lower or particularly upper) in their copula structure?
and (Q2) if they do show such asymmetric tail-dependence (this is a property of 
non-Normal copula structure) then what causes such tail dependence? We hypothesized 
possible biotic and/ or abiotic drivers causes such inter-specific synchrony.


# Materials and Methods\label{M&M}

## Data \label{Data} 

Our first spatio-temporal dataset comprised average annual 
abundance estimates for 26 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 plankton taxa for the 
56 years from 1958 to 2013 [@sheppard2017]. 
The locations are $2^\circ$ by $2^\circ$ grid cells.
These data were obtained from the Continuous Plankton Recorder (CPR) dataset, 
now operated and maintained by 
the Marine Biological Association of the United Kingdom. 
Data preprocessing steps are same as used in Ghosh et al. (in review). 
As dinoflagellate are considered having important role in phytoplankton bloom,
we screened 4 Ceratium species (Table \ref{tab_plankton_aphid_info}) and 
their generalist predator zooplankton \textit{Calanus finmarchicus} (hereafter abbreviated as C. fin.)
from 15 selected locations. Screening of locations and species involved several steps: 
first, we chose a subset, from 26 locations, which had atleast 36 years of data available for all species.
Second, for a given location, we screened species excluding the ones that were absent for more 
than 10$\%$ of sampling years at that location. Finally, 
we considered only those selected locations which had atleast any two 
\textit{Ceratium} sp. along with their predator C. fin. We also used
time series of temperature (for growing season) at each selected grid (Fig. \ref{SM-fig_plankton_map}) 
to be one of the key drivers
which caused synchronous response among \textit{Ceratium} sp.

As second spatio-temporal dataset, we used phenological data of aphid first-flight dates for 10 selected 
among total of 11 sites (Fig. \ref{SM-fig_aphid_map}) across the United Kingdom (UK). 
Data were available for 20 aphid species (Table \ref{tab_plankton_aphid_info}). Locations are screened 
based on atleast 30 years of data points available for all sp. at each site 
during the timespan of 1976-2010. These time series were computed 
from the the Rothamsted Insect Survey suction-trap dataset 
[@harrington2014; @bell2015]. Data pre-processing for the aphids
was the same as that of [@sheppard2016]. Time-series of winter average 
temperature (December, $n^{th}$ year to March, $(n+1)^{th}$ year) was considered as 
main driver causing synchronous arrival of aphids at each site.

## Methods\label{Methods}

To begin with, we considered $T$ years long time series were available for $s$ number of species along with 
$d$ abiotic and/ or biotic drivers (i. e. temperature or predator abundance) for any $n^{th}$ site.
For each selected location we constructed a bivariate copula made of time series of either
two species or with a species and time series of the driver
in the following way: given a bivariate dataset from $n^{th}$ site, $(x_t^n,y_t^n)$ for $t=1,...,T$, 
first, we produced ranks $(\tilde{u}_t^n,\tilde{v}_t^n)$ where $\tilde{u}_t^n$ is the rank of $x_t^n$ in the set 
$\{x_t^n:t=1,...,T\}$ and $\tilde{v}_t^n$ is the rank of $y_t^n$ in the set $\{y_t^n:t=1,...,T\}$. 
we set $u_t^n=\tilde{u}_t^n/(T+1)$ and $v_t^n=\tilde{v}_t^n/(T+1)$; subsequent analysis used these 
*normalized ranks* $u_t^n \in [0,1]$ and $v_t^n \in [0,1]$. Ranks were $1$ for the smallest element in a set. 
Second, we did a test of the independence of the $u_t^n$ and $v_t^n$ 
(implemented in `BiCopIndTest` in the `VineCopula` package). 
If independence could be rejected ($0.05$ significance level), and $(u_t^n, v_t^n)$
were postively correlated then we proceed to measure dependence. For significantly negative 
association, we started with $(x_t^n, - y_t^n)$ so that $(u_t^n, v_t^n)$ remained positively correlated.
we computed the partial Spearman correlation statistics $\cor_{l_b,u_b}$ as introduced in Ghosh et al. (in review)
as dependence measure for a positively correlated bivariate copula. 

Given lower and upper boundaries by two parallel lines ($u_t^n+v_t^n = 2l_b$, $u_t^n+v_t^n = 2u_b$ such that $0 \leq l_b < u_b \leq 1$) 
which scans the region of interest of the copula within a unit square box, we here
redefine the statistics we computed within the bounds; 
\begin{equation}\label{eq.Cor}
   \cor_{l_b,u_b}^n(u,v) = \frac{\sum 
   (u_t^n-\mean(u)) (v_t^n-\mean(v))}{(T-1)\sqrt{\var(u)\var(v)}}.
\end{equation}
Here means and sample variances are computed using all $T$ data points for $n^{th}$ location, but the
sum is over only the indices $t$ for which $u_t^n+v_t^n > 2l_b$ and $u_t^n+v_t^n < 2u_b$. 
The sum of $\cor_{0,1/2}^n(u,v)$ and $\cor_{1/2,1}^n(u,v)$ (or some other choice of 
$\cor_{l_{b_k},u_{b_k}}^n(u,v)$ for bounds $l_{b_k},u_{b_k}$ that partition the 
interval $(0,1)$) equals the standard Spearman correlation.

<!--## Methods to answer (Q1)-->
If a bivariate copula was made with *normalized ranks* $(u, v)$ of any two time series 
of species pair $(i, j)$ taken from $n^{th}$ site, then we used the shorthand notation
for lower tail dependence as $C^n_L(i, j) = \cor_{0,b}^n(u,v)$ 
and upper tail dependence as $C^n_U(i, j) = \cor_{1-b,1}^n(u,v)$ where $0 \leq b \leq 0.5$,
respctively. We considered $b = 1/3$ for plankton data, whereas due to fewer data points we used 
$b = 1/2$ for aphids. Higher partial Spearman 
correlation in lower tail, i.e., when $C^n_{L-U}(i, j) = C_L^n(i, j)  - C_U^n(i, j)$ is positive, 
indicates asymmetry in tail-dependence of a copula and to be specific 
stronger lower tail dependence than upper. Similarly, negative value of 
$C^n_{L-U}(i, j)$ would indicate stronger upper tail 
dependence than lower in an asymmetric copula. 
For $n^{th}$ site, we estimated $C^n_{L-U}(i, j)$ for all pairwise combination of $(i, j)$ and counted
number of lower ($N_L$) and upper ($N_U$) tail dominant species pair for which 
$C^n_{L-U}(i, j)>0$ and $C^n_{L-U}(i, j)<0$, respectively.
We also added up all positive and negative values of $C^n_{L-U}(i,j)$ and defined lower and upper
asymmetric estimates on species level for $n^{th}$ site as $A_S^{n+} = \sum_{i=1}^{s}\sum_{j=1}^{s}[C^n_{L-U}(i, j) > 0]$ and 
$A_S^{n-} = \sum_{i=1}^{s}\sum_{j=1}^{s}[C^n_{L-U}(i, j) < 0]$, 
respectively. Total species-asymmetry $A_S^{n} = A_S^{n+} + A_S^{n-}$ was assessed for 
each selected site which could
answer (Q1); if $A_S^{n} \neq 0$ then it would indicate asymmetry in 
overall tail-dependence. But to get more precise answer, whether the asymmetric
tail dependence was stronger in lower values or upper, proportion 
of lower and upper tail dominance
in overall asymmetry at each site were measured
by $f_L = \frac{A_S^{n+}}{A_S^{n+} + |A_S^{n-}|}$ and
$f_U = \frac{|A_S^{n-}|}{A_S^{n+} + |A_S^{n-}|}$, respectively.
These values of $f_L$ and $f_U$ from each site were used to answer (Q1). 

<!--## Methods to answer (Q2)-->

To answer (Q2), our approach was to investigate the causal relationship between 
inter-species tail dependence and tail-dependence between
species with biotic and/ or abiotic drivers. We used two slightly different estimate to answer (Q2) 
for our two datasets. First, we estimated lower minus upper tail dependence
$C_{L-U}^n(i, k); i = 1\dots s, k = 1\dots d$ 
for the copulas made with species and drivers' time series from any $n^{th}$ site
as we did with copulas for any two species pair combination. Then, we measured total asymmetry 
on drivers level as $A_D^n = \sum_{i=1}^{s}\sum_{k=1}^{d} C^n_{L-U}(i, k)$ and tested if there was
any significant correlation between $A_D^n$ and $A_S^n$ considered for all selected sites. This 
approach was better when we examined behaviour of fewer but similar species, like the four Ceratium sp. 
of dinoflagellate class. Our second approach was more suitable when 
there were several species but not from same genus, so expected to show more variation
in inter-species response, as we will see in 20 aphid species' firstflight response. 
Here we examined correlation between response for $i^{th}$ individual species 
with all the other species ($\alpha_S^n(i)$) and with its drivers ($\alpha_D^n(i)$). 
We measured those two quantities as $\alpha_S^n(i) = \sum_{j=1}^{s} C^n_{L-U}(i, j)$ and 
$\alpha_D^n(i) = \sum_{k=1}^{d} C^n_{L-U}(i, k)$ for a given $i^{th}$ species. Significant correlation 
between $A_D^n$ and $A_S^n$ (for planktons) or between $\alpha_D^n(i)$ and $\alpha_S^n(i)$ (for aphids)
would establish our hypothesis that tail dependence in inter-specific response was caused by similar
tail-dependence in species-driver copulas.


# Results\label{Results}

<!--Results for Q1-->
We found major asymmetric tail-dependence when we considered all pairwise combination of any 
two species time-series from a given location.
Summarized results for all selected locations showing asymmetry in either tail for \textit{Ceratium} sp. 
was depicted in Fig. \ref{fig_CorlmCoru_plankton_map_loc12_26}(A), where we plotted $f_L$ and
$f_U$ (see \nameref{Methods}) for each selected sites 
(except site 3 as there were no significant dependence between species abundances). Site
codes in map were mentioned in red color if sites showed 
stronger lower tail dependence contribution (i.e. $f_L > f_U$ for synchronous rare 
occurrence of \textit{Ceratium}s). Similarly, locations where \textit{Ceratium} sp. were synchronously 
abundant (site codes in blue color of Fig. \ref{fig_CorlmCoru_plankton_map_loc12_26}A), had 
more upper tail dependent contribution ($f_U > f_L$). Partial Spearman correlation-based lower minus upper 
tail dependence matrix ($C_{L-U}^n$) for site $n=12$ (Fig. \ref{fig_CorlmCoru_plankton_map_loc12_26}B) and $n=26$ (Fig. \ref{fig_CorlmCoru_plankton_map_loc12_26}C) were shown as examplary case (for 
other locations see Fig. \ref{SM-fig_CorlmCoru_plankton_all_loc}). 
Whereas Location 12 showed dominantly lower tail-dependent interspecific synchrony with all 12 pairwise copula had 
lower tail dependence ($N_L = 12, N_U = 0$), Location 26 showed stronger upper tail dependence
($N_L = 2, N_U = 4$). Similar conclusion was made for aphid firstflight time series, where copulas showed
stronger tail-dependence in either tail (lower or upper) for all selected sites 
(Fig. \ref{fig_CorlmCoru_ff_map_loc6_7}). Major asymmetric tail-dependence observed with one population (planktons)
and one phenological (aphids firstflight) data answered (Q1) that inter-specific synchrony was captured through tail-dependence
estimates of copulas from community matrix or sp-sp interaction matrix, $C_{L-U}^n$ for a given location $n$.

<!--plankton plots : Q1-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=10 cm]{./Results/plankton_results/npa_stat_results/Corstat_LmU_values_on_map_sp_only.pdf} 
\includegraphics[width=8 cm]{./Results/plankton_results/npa_stat_results/loc12/loc12_Corl-Coru.pdf}
\includegraphics[width=8 cm]{./Results/plankton_results/npa_stat_results/loc26/loc26_Corl-Coru.pdf}
\caption{$(A)$ Proportion of tail dominant contribution ($f_L$ for lower tail and $f_U$ for upper tail 
showing on a color scale between 0 to 1) in overall asymmetry, plotted for each 
selected locations among 26 sites (see Fig. \ref{SM-fig_plankton_map}) separately for sp-sp interaction matrix,
summarizes the overall 
asymmetry in tail-dependence for all sampling sites. Site 3 has independent relationship between species. 
Lower tail - upper tail dependence matrix ($C_{L-U}^n$ for location n = 12 and n = 26) with Ceratium abundances 
are showin in $(B)$ and $(C)$, respectively. For other locations see Fig. \ref{SM-fig_CorlmCoru_plankton_all_loc}.
The last 2 rows and columns are for the estimates from interaction between Ceratium sp. and  
drivers (\textit{Calanus finmarchicus}, growing season temperature).
$N_L$ and $N_U$ are the count of cells with positive and 
negative values for positively correlated species pair. $N_L > N_U$ in $(B)$ 
indicates major lower tail dependence in sp-sp interaction matrix. Similarly, $N_U > N_L$ in $(C)$ indicates more upper
tail dependence. Yellow cells indicate independent relationship. Cells with green dot indicates associated 
variables were negatively correlated initially and hence we reversed the variables in column so that the copula had positive association.\label{fig_CorlmCoru_plankton_map_loc12_26}}
\end{center}
\end{figure}

<!--aphid plots : Q1-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=10 cm]{./Results/aphid_results/ff_npa_stat_results/Corstat_LmU_values_on_map_sp_only.pdf} 
\includegraphics[width=8 cm]{./Results/aphid_results/ff_npa_stat_results/loc6/loc6_Corl-Coru.pdf}
\includegraphics[width=8 cm]{./Results/aphid_results/ff_npa_stat_results/loc7/loc7_Corl-Coru.pdf}
\caption{$(A)$ Proportion of tail dominant contribution ($f_L$ for lower tail and $f_U$ for upper tail 
showing on a color scale between 0 to 1) in overall asymmetry were plotted for each 
selected 10 locations (see Fig. \ref{SM-fig_aphid_map}) separately for 20 $\times$ 20 sp-sp 
interaction matrix summarizes the overall 
asymmetry in tail-dependence for all sampling sites. Lower tail - upper tail dependence matrix 
($C_{L-U}^n$ for location n = 6 and n = 7) with first flight
data of 20 aphid species (for other locations see Fig. \ref{SM-fig_CorlmCoru_ff_all_loc}.), 
the last row and column indicates estimates from interaction between each 
aphid species and winter average temperature (which was supposed to be the driver of the species' synchrony),
details were discussed in \nameref{Results}. $N_L$ and $N_U$ are the count of positive and 
negative cells for 20$\times$ 20 matrix; $N_L > N_U$ in $(B)$ 
indicates major lower tail dependence in sp-sp interaction matrix. Similarly, $N_U > N_L$ in $(C)$ indicates more upper
tail dependence. Yellow cells indicate independent relationship. Cells with green dot indicates associated 
variables were negatively correlated initially and hence we reversed the variables in column so that the copula had positive association.\label{fig_CorlmCoru_ff_map_loc6_7}}
\end{center}
\end{figure}


<!--Results for Q2-->





<!--plankton scatter plot : Q2-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=10 cm]{./Results/plankton_results/npa_stat_results/Corstat_scatter_LmU_values.pdf} 
\caption{plghjjjjjjjjjjjjjj\label{fig_plankton_scatter}}
\end{center}
\end{figure}


<!--aphid multipanel plot : Q2-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/aphid_results/ff_npa_stat_results/Corstat_LmU_avg_sp_temp_multipanel_plot.pdf} 
\caption{plghjjjjjjjjjjjjjj\label{fig_aphid_multipanel}}
\end{center}
\end{figure}





# Discussion\label{Discussion}




<!--Table with species info -->
```{r tab_plankton_aphid_info, echo=F, results='asis',message=F}
library(tinytex)
library(tibble)
library(kableExtra)
library(dplyr)

#aphid_info_org <- tibble(c0=c(1:20),
#                     c1=c("Apple grass aphid","Bird cherry oat aphid","Black bean aphid","Blackberry cereal aphid",
#                       "Blackcurrant sowthistle aphid",
#                      "Corn leaf aphid","Currant lettuce aphid","Damson hop aphid","Grain aphid","Green spruce aphid",
#                      "Leaf-curling plum aphid","Mealy cabbage aphid","Mealy plum aphid","Pea aphid","Peach potato aphid",
#                      "Potato aphid","Rose grain aphid","Shallot aphid","Sycamore aphid","Willow carrot aphid"),
#                    c2=c("Rhopalosiphum insertum", "Rhopalosiphum padi", "Aphis fabae", "Sitobion fragariae", "Hyperomyzus lactucae",
#                      "Rhopalosiphum maidis", "Nasonovia ribisnigri","Phorodon humuli","Sitobion avenae","Elatobium abietinum",
#                       "Brachycaudus helichrysi","Brevicoryne brassicae","Hyalopterus pruni","Acyrthosiphon pisum","Myzus persicae",
#                       "Macrosiphum euphorbiae","Metopolophium dirhodum","Myzus ascalonicus","Drepanosiphum platanoidis","Cavariella #aegopodii")
#)

plankton_aphid_info <- tibble(c0=c(1:4,1:20),
                     c2=c("Ceratium fusus","Ceratium furca","Ceratium tripos","Ceratium macroceros",
                          "Rhopalosiphum insertum", "Rhopalosiphum padi", "Aphis fabae", "Sitobion fragariae",
                          "Hyperomyzus lactucae", "Rhopalosiphum maidis", "Nasonovia ribisnigri","Phorodon humuli",
                          "Sitobion avenae","Elatobium abietinum", "Brachycaudus helichrysi","Brevicoryne brassicae",
                          "Hyalopterus pruni","Acyrthosiphon pisum","Myzus persicae", "Macrosiphum euphorbiae",
                          "Metopolophium dirhodum","Myzus ascalonicus","Drepanosiphum platanoidis","Cavariella aegopodii"                                           ))
knitr::kable(plankton_aphid_info, "latex", booktabs = T, linesep = "",align="c",
      caption = "Species names of 4 planktons and 20 aphids for which data were used \\label{tab_plankton_aphid_info}",
      col.names = NULL)%>%
      group_rows("Planktons", 1,4) %>%
      group_rows("Aphids", 5,24) %>%
      column_spec(column = 2,italic=T)%>%add_header_above(header=c("Species id" = 1, "Latin binomial" = 1))
```













\clearpage

# Literature cited

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{1pt}
\noindent
